### 2015-05-13 whoop.tex
Addressed Pantazis's comment by deleting the associated sentence :-)

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race in a shared variable. We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analysing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.  The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts that the lockset for each shared variable is non-empty.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.

@@

\medskip\noindent\textbf{Concurrent Programming Model }
%
We consider a simple concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in Section~\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.  A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.  A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.  A procedure in $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.  A basic block either has a single successor or a pair of successors.  In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.

@@

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.

At any point of execution, a new thread may be added to the list of thread states.  This involves selecting an id $i \in \ids$ that has not been previously used during program execution and a thread template $T$, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  The thread's next statement is updated \PDComment{Do you mean execute?} according to the control flow graph of the current procedure.

@@

Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.

>>>>

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analysing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.  The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.

@@

\medskip\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in Section~\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.  A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.  A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of $T$ by a thread.  Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.  A basic block either has a single successor or a pair of successors.  In the latter case, an \emph{exit condition} over thread-private variables determines the successor to which control should flow on block exit.

@@

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.

@@

\ADComment{Renaming must talk about basic block exit conditions} Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.
